Elaboration with First-Class Implicit Function Types

### 2020-02-26 paper.tex
stuff

Our telescopes and records are derivable from natural numbers, the unit type and
$\Sigma$-types. We use Agda-like pattern matching notation in the following.
First, we define length-indexed telescopes.
\begin{alignat*}{3}
  & \rlap{$\Tel' : \Nat \to \U$}\\
  & \Tel'\,\zero      && :\equiv \top\\
  & \Tel'\,(\suc\,n)  && :\equiv \Sigma(A : \U).\,(A \to \Tel'\,n)
\end{alignat*}
Then, we have $\Tel :\equiv \Sigma(n : \Nat).\,(\Tel'\,n)$, and define records:
\begin{alignat*}{3}
  & \rlap{$\Rec : \Tel \to \U$}\\
  & \Rec\,(\zero,\,\_)        &&:\equiv \top\\
  & \Rec\,(\suc\,n,\,(A,\,B)) &&:\equiv \Sigma(a : A).\,(\Rec\,(n,\,B\,a))
\end{alignat*}
From the above, $\epsilon$, $\blank\TCons\blank$, $\blank::\blank$ and ${[}{]}$
are evident, and all expected equalities hold definitionally. Derivability is
good news because we inherit nice properties of the type theory which only
contains the base type formers. For instance, if we have a consistent universe
setup\footnote{Recall that we currently assume type-in-type, which causes consistency and normalization to fail.}, we inherit consistency, canonicity and decidability of conversion. We
currently use native telescopes instead of $\Nat$, $\top$ and $\Sigma$ because
in unification and elaboration it is convenient that we are able to restrict
some types to records types.

@@

\section{Evaluation}
\label{sec:impl_and_eval}

@@

\begin{figure}
\begin{tabular}{|lll|}
\hline
A & \multicolumn{2}{l|}{\textsc{polymorphic instantiation}}\\
\hline
A1 & $\lambda\,x\,y.\,y$ & \yesst\\
A2 & $choose\,id$ & \yesst \\
A3 & $choose\,nil\,ids$ & \yes \\
A4 & $\lambda\,(x : \{A : \U\}\to A \to A).\,x\,x$ & \yesst\\
A5 & $id\,auto$ & \yes\\
A6 & $id\,auto'$ & \yesst\\
A7 & $choose\,id\,auto$ & \yes\\
A8 & $choose\,id\,auto'$ & \no\\
A9 & $\lambda\,(f : \{A : \U\}\to(A\to A)\to List\,A \to A).\,f\,(choose\,id)\,ids$ & \yesst\\
A10 & $poly\,id$ & \yes\\
A11 & $poly\,(\lambda\,x.\,x)$ & \yes\\
\hline
B & \multicolumn{2}{l|}{\textsc{inference of polymorphic arguments}}\\
\hline
B1 & $\lambda\,f.\,pair\,(f\,zero)\,(f\,true)$ & \no\\
B2 & $\lambda\,xs.\,poly\,(head\,xs)$ & \no\\
\hline
C & \multicolumn{2}{l|}{\textsc{functions on polymorphic lists}}\\
\hline
C1  & $length\,ids$ & \yes \\
C2  & $tail\,ids$ & \yes \\
C3  & $head\,ids$ & \yesst \\
C4  & $single\,id$ & \yesst \\
C5  & $cons\,id\,ids$ & \yes \\
C6  & $cons\,(\lambda\,x.\,x)\,ids$ & \yes \\
C7  & $append\,(single\,inc)\,(single\,id)$ & \yes \\
C8  & $\lambda\,(g:\{A:\U\}\to List\,A \to List\,A \to A).\,g\,(single\,id)\,ids$ & \yesst \\
C9  & $map\,poly\,(single\,id)$ & \yes \\
C10 & $map\,head\,(single\,ids)$ & \yes \\
\hline
D & \multicolumn{2}{l|}{\textsc{application functions}}\\
\hline
D1 & $app\,poly\,id$        & \yes \\
D2 & $revapp\,id\,poly$     & \yes \\
D3 & $runST\,argST$         & \yes \\
D4 & $app\,runST\,argST$    & \yes \\
D5 & $revapp\,argST\,runST$ & \yes \\
\hline
E & \multicolumn{2}{l|}{\textsc{$\eta$-expansion}}\\
\hline
& \multicolumn{2}{l|}{assuming $k : \{A : \U\}\to A \to List\,A \to A$,} \\
& \multicolumn{2}{l|}{$h : Int \to \{A:\U\}\to A \to A$}\\
& \multicolumn{2}{l|}{and $lst : List\,(\{A:\U\} \to Int \to A \to A)$} \\
E1 & $k\,h\,lst$ & \no \\
E2 & $k\,(\lambda\,x.\,h\,x)\,lst$ & \yesst \\
E3 & $\lambda\,(r : (\{A :\U\}\to A \to \{B : \U\}\to B \to B) \to Int).\,r\,(\lambda\,x\,y.\,y)$ & \yes \\
\hline
\multicolumn{3}{l}{\text{{\footnotesize
``Yes*'' means that our system can infer a type for the
expression, but the lack of}}}\\
\multicolumn{3}{l}{\text{{\footnotesize
let-generalization causes unsolved metas in the type.}}}
\end{tabular}

\caption{Elaboration benchmark from \cite{serrano2018guarded}.}
\end{figure}

}

%% (h : Int → {A} → A → A)(k : {A} → A → List A → A)(lst : List ({A} → Int → A → A))

>>>>

Telescopes and records are derivable from natural numbers, the unit type and
$\Sigma$-types. We use Agda-like pattern matching notation in the following.
First, we define length-indexed telescopes.
\begin{alignat*}{3}
  & \rlap{$\Tel' : \Nat \to \U$}\\
  & \Tel'\,\zero      && :\equiv \top\\
  & \Tel'\,(\suc\,n)  && :\equiv \Sigma(A : \U).\,(A \to \Tel'\,n)
\end{alignat*}
Then, we have $\Tel :\equiv \Sigma(n : \Nat).\,(\Tel'\,n)$, and define records:
\begin{alignat*}{3}
  & \rlap{$\Rec : \Tel \to \U$}\\
  & \Rec\,(\zero,\,\_)        &&:\equiv \top\\
  & \Rec\,(\suc\,n,\,(A,\,B)) &&:\equiv \Sigma(a : A).\,(\Rec\,(n,\,B\,a))
\end{alignat*}
From the above, $\epsilon$, $\blank\TCons\blank$, $\blank::\blank$ and ${[}{]}$
are evident, and all expected equalities hold definitionally. Derivability is
good news because we inherit nice properties of the type theory which only
contains the base type formers. For instance, if we have a consistent universe
setup\footnote{Recall that we currently assume type-in-type, which causes consistency and normalization to fail.}, we inherit consistency, canonicity and decidability of conversion. We
currently use native telescopes instead of $\Nat$, $\top$ and $\Sigma$ because
in unification and elaboration it is convenient that we are able to restrict
some types to records types.

@@

\section{Related Works and Evaluation}
\label{sec:impl_and_eval}

In this section we examine how the elaborator fares in practice, its
limitations, and how it compares to related works. We abbreviate our elaborator
as FCIF when comparing it to others.

\subsubsection{Related works}
MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
bounds, which supports strong inference for first class polymorphism. HML
\cite{leijen2009flexible}) is a simplified variant of MLF. Since these systems
diverge markedly from (fragments of) Martin-Löf type theory and rely critically
on subtyping, it is unclear how they could be extended to dependent type
theories. HMF \cite{leijen2008hmf} is a relatively simple system with
first-class polymorphism; however, it is also weaker than others in terms of
inferable annotations, and its more powerful variant (extended with n-ary
application handling) jumps in complexity.

There have been several attempts in the context of GHC. Boxy types
\cite{vytiniotis2006boxy} and FPH \cite{vytiniotis2008fph} were two early
iterations which suffered from complex specification, complex implementation or
fragility. More recent works are guarded impredicativity (GI)
\cite{serrano2018guarded} and quick look impredicativity (QL)
\cite{serrano2020a}. They both try to examine n-ary applications to find
metavariable instantiations which arise from occurrences guarded by rigid type
constructors. GI's use of \emph{generalization constraints} anticipates our use
of telescopes, but overall GI is also burdened by inessential complexity. QL
streamlines GI by eschewing constraints in favor of an eager preprocessing pass
on neutral expressions which finds polymorphic instantiations. QL also takes
advantage of bidirectional type propagation.  Although QL seems to be
practically the most favorable so far, it is far from being elegant; it does
multiple checking passes, and the restriction of quick looking to neutral
applications is rather ad-hoc.

@@

\begin{figure}
\begin{alignat*}{4}
  & IdTy   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && inc    && : Int \to Int                                        \\
  & single && : \{A : \U\} \to A \to List\,A                       && auto   && : IdTy \to IdTy                                      \\
  & id     && : IdTy                                               && auto'  && : \{B : \U\}\to IdTy \to B \to B                     \\
  & ids    && : List\,IdTy                                         && choose && : \{A : \U\} \to A \to A \to A                       \\
  & nil    && : \{A : \U\}\to List\,A                              && app    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
  & cons   && : \{A : \U\} \to A \to List\,A \to List\,A           && revapp && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
  & head   && : \{A : \U\} List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
  & tail   && : \{A : \U\} List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
  & map    && : \{A B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool
\end{alignat*}
\caption{Types used in Figure \ref{fig:benchmark}.}
\label{fig:benchmark_types}
\end{figure}

\begin{figure}
\begin{tabular}{|lll|}
\hline
A & \multicolumn{2}{l|}{\textsc{polymorphic instantiation}}\\
\hline
A1 & $\lambda\,x\,y.\,y$ & \yesst\\
   & \text{we infer a type which may be solved to $\{A\,B : \U\}\to A \to B \to A$} &\\
   & \text{but not to $\{A : \U\}\to A \to \{B : \U\} \to B \to A$} & \\
A2 & $choose\,id$ & \yesst \\
A3 & $choose\,nil\,ids$ & \yes \\
A4 & $\lambda\,(x : \{A : \U\}\to A \to A).\,x\,x$ & \yesst\\
A5 & $id\,auto$ & \yes\\
A6 & $id\,auto'$ & \yesst\\
A7 & $choose\,id\,auto$ & \yes\\
A8 & $choose\,id\,auto'$ & \no\\
A9 & $\lambda\,(f : \{A : \U\}\to(A\to A)\to List\,A \to A).\,f\,(choose\,id)\,ids$ & \yesst\\
A10 & $poly\,id$ & \yes\\
A11 & $poly\,(\lambda\,x.\,x)$ & \yes\\
\hline
B & \multicolumn{2}{l|}{\textsc{inference of polymorphic arguments}}\\
\hline
B1 & $\lambda\,f.\,pair\,(f\,zero)\,(f\,true)$ & \no\\
B2 & $\lambda\,xs.\,poly\,(head\,xs)$ & \no\\
\hline
C & \multicolumn{2}{l|}{\textsc{functions on polymorphic lists}}\\
\hline
C1  & $length\,ids$ & \yes \\
C2  & $tail\,ids$ & \yes \\
C3  & $head\,ids$ & \yesst \\
C4  & $single\,id$ & \yesst \\
C5  & $cons\,id\,ids$ & \yes \\
C6  & $cons\,(\lambda\,x.\,x)\,ids$ & \yes \\
C7  & $append\,(single\,inc)\,(single\,id)$ & \yes \\
C8  & $\lambda\,(g:\{A:\U\}\to List\,A \to List\,A \to A).\,g\,(single\,id)\,ids$ & \yesst \\
C9  & $map\,poly\,(single\,id)$ & \yes \\
C10 & $map\,head\,(single\,ids)$ & \yes \\
\hline
D & \multicolumn{2}{l|}{\textsc{application functions}}\\
\hline
D1 & $app\,poly\,id$        & \yes \\
D2 & $revapp\,id\,poly$     & \yes \\
D3 & $runST\,argST$         & \yes \\
D4 & $app\,runST\,argST$    & \yes \\
D5 & $revapp\,argST\,runST$ & \yes \\
\hline
E & \multicolumn{2}{l|}{\textsc{$\eta$-expansion}}\\
\hline
& \multicolumn{2}{l|}{assuming $k : \{A : \U\}\to A \to List\,A \to A$,} \\
& \multicolumn{2}{l|}{$h : Int \to \{A:\U\}\to A \to A$}\\
& \multicolumn{2}{l|}{and $lst : List\,(\{A:\U\} \to Int \to A \to A)$} \\
E1 & $k\,h\,lst$ & \no \\
E2 & $k\,(\lambda\,x.\,h\,x)\,lst$ & \yesst \\
E3 & $\lambda\,(r : (\{A :\U\}\to A \to \{B : \U\}\to B \to B) \to Int).\,r\,(\lambda\,x\,y.\,y)$ & \yes \\
\hline
\multicolumn{3}{l}{\text{{\footnotesize
``Yes*'' means that our system can infer a type for the
expression, but the lack of}}}\\
\multicolumn{3}{l}{\text{{\footnotesize
let-generalization yields unsolved metas in the type.}}}
\end{tabular}

\caption{Elaboration benchmark from \cite{serrano2018guarded}.}
\label{fig:benchmark}
\end{figure}
}

\subsubsection{Evaluation}
We borrow a very useful collection of inference benchmarks from the GI
\cite{serrano2018guarded} and QL \cite{serrano2020a} papers: see Figure
\ref{fig:benchmark_types} and Figure \ref{fig:benchmark}. We also include a
source file in the implementation which reproduces these results.  In ibid.\ a
comparison is presented between multiple systems, but here we only include
FCIF. The relative performance of FCIC here is easy to remember: it handles
exactly the same cases as QL (which is slightly more than what GI covers). We
mark some cases as ``Yes*'', where FCIF successfully infers a type, but since
FCIC does no let-generalization, the inferred types are not fully constrained
without extra contextual information.

In the $A1$ case, FCIF inserts only a single curried $\lambda$ on the outside,
which is why the inferred type is not unifiable with $\{A : \U\}\to A \to \{B :
\U\} \to B \to A$. However, this could be easily remedied by adding an extra
curried $\lambda$ insertion in the definition of
$\einfer{\lambda\,x.\,t}{ins}{\Theta}{\Gamma}$.

In $A8$, the failure is intentional: the types of $id$ and $auto'$ are not
unifiable because of the mismatched order of implicit and explicit arguments. We
do not float out or reorder implicit arguments in any way, because we want to
support arbitrary mixing of implicit/explicit arguments in the surface language,
and such reordering would be problematic anyway in the presence of dependent
types. The same situation arises in $E1$, where we cannot unify $\{A : \U\}\to
Int \to A \to A$ with $Int \to \{A : \U\}\to A \to A$.

While in this benchmark FCIF matches QL, in other ways it is significantly
stronger. Some general observations on the relative strength of FCIF:
\begin{itemize}
  \item
        Most obviously, FCIF supports dependent type theory and prior solutions
        do not. It is also doubtful whether prior solutions scale to dependent
        types. In particular, QL's preprocessing step is not applicable, since in
        a dependently typed spine we cannot do anything with later arguments until
        previous arguments are actually elaborated, or else it has to be restricted
        to non-dependent spines.
  \item
        FCIF is global and does not consider neutral spines specially in any
        way. For an example for global inference, $\slet\,x : \_ =
        single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF but does not
        work in GI, QL and HMF. It also works in MLF and HML, which are also
        global in this sense. These work by immediately giving a principal type
        to $single\,id$ which includes a subtyping bound. In contrast, FCIF
        makes no promises about principal typing --- which is not feasible with
        dependent types --- but it can still infer a type for $id$ which can be
        later constrained to $IdTy$.
\end{itemize}

\emph{Limitations.} What FCIF definitely cannot do, is inferring an implicit
function type for a function argument. MLF is the only system which can do this,
in cases where the polymorphic argument is used only once. Consider the
following:
\[
  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. In the next
section we sketch an extension of FCIF which can handle this, and can also
handle $B2$ from Figure \ref{fig:benchmark}.

\section{Future Work}

\section{Conclusion}

%% -- let f : IdTy → Bool = λ x. true in
%% -- λ x. f x
